\documentclass[11pt]{article}
\usepackage[preprint]{acl}
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage{array}
\usepackage{graphicx}
\usepackage{tabularx}
\setcitestyle{numbers}
\setlength\titlebox{6.5cm}
\title{Understanding LLM-MAS Before Shipping It:\\A Beginner-to-Advanced Survey (2024--2026Q1)}
\author{Draft Manuscript\\
  Multi-generative\_Agent\_System\_survey}

\begin{document}
\maketitle

\begin{abstract}
Large-language-model-based multi-agent systems (LLM-MAS) combine models, tools, state, and sometimes multiple roles to solve multi-step tasks. Updating a 2024 baseline with 2025--2026Q1 literature and industry signals, this survey covers concepts, architecture, training, evaluation, failure modes, and deployment. Our central claim is that recent progress comes mainly from runtime maturation: orchestration, state, policy enforcement, observability, evaluation, and safety.
\end{abstract}

\section{Introduction}
LLM-MAS is used loosely. This survey uses an operational definition rather than the classical distributed-MAS definition. We focus on LLM-centered systems that execute multi-step trajectories through tools, state, and explicit control flow; the multi-agent case is the stricter subset with operationally separated roles. Prompt-only personas and generic distributed decision systems without an LLM-centered runtime are outside scope. The beginner problem is taxonomy: what counts as agentic, what counts as multi-agent, and why strong demos fail to transfer.

This survey therefore starts with concepts, system components, training, evaluation, and limitations before turning to research and deployment. For a newcomer, mechanism comes before shipping judgment.

\paragraph{Scope.}
We focus on 2025 and 2026Q1 while retaining enough 2023--2024 context to anchor the shift. We de-emphasize role-play demos unless they contribute reusable benchmarks, datasets, realism metrics, or alignment methods.

\paragraph{Thesis.}
The field is converging on a systems view: progress depends less on prompt design than on the end-to-end runtime stack.

\paragraph{2025--2026Q1 shift.}
The center of gravity moved for three reasons: high-value tasks became tool-bound, static benchmark gains stopped being enough for deployment judgment, and autonomy expanded the risk surface. As a result, runtime control, replayability, policy-mediated execution, and regression gates became more important than prompt novelty alone.

\paragraph{Judgment boundary.}
This survey treats role-play or debate setups without reusable data, behavioral metrics, or verification loops as weak evidence. It likewise discounts automation stacks that lack permission boundaries, replayability, or failure diagnostics.

\section{Architecture, Coordination, and Interfaces}
For current LLM-MAS, the useful abstraction is a runtime stack rather than a set of abstract agents in an environment. The main design question is how to place orchestration, execution, checking, state, and control boundaries in a trajectory-producing runtime. Multi-agent structure is justified only when it yields decomposition, independent checking, or parallel execution; otherwise one orchestrated model plus tools and typed state is usually the cleaner description.

\subsection{Architecture and coordination patterns}
\textbf{Centralized orchestration} is the deployable default because policy, audit, and rollback are easier to enforce from one runtime. \textbf{Decentralized coordination} should be reserved for cases where agents truly hold different tools or information and the extra debugging cost is justified.

\subsubsection{Planner--Executor--Verifier pipelines}
A common high-value pattern separates concerns:
\begin{itemize}[leftmargin=1.2em]
  \item planner: decomposition and sequencing,
  \item executor: tool actions and intermediate artifacts,
  \item verifier/reviewer: constraint checks and correctness gates.
\end{itemize}
This decomposition is useful only if intermediate artifacts are inspectable and verifier independence is real; otherwise the split becomes theater rather than reliability.

\subsubsection{Hierarchical teams and graph workflows}
Hierarchical teams help with deep decomposition but require stronger termination rules and state discipline. Graph orchestration matters because it makes plan, execution, review, retry, and rollback explicit and controllable.

\subsubsection{Failure taxonomy}
The most common failure modes are:
\begin{itemize}[leftmargin=1.2em]
  \item planning drift,
  \item tool thrashing,
  \item message overrun,
  \item hidden coupling,
  \item verification collapse,
  \item premature consensus,
  \item role inconsistency.
\end{itemize}
Typed role contracts, structured state passing, review gates, and deterministic checkpoints usually help more than adding more conversational turns.

\subsection{Protocols as an interface layer}
Protocols matter mainly as an interface layer between models, tools, data sources, and sometimes other agent runtimes. The details are moving quickly, so this survey does not treat any one protocol design as a stable technical center of the field. The durable point is simpler: protocolization reduces bespoke integration work, but it also turns identity, permissions, audit, and trust boundaries into first-class runtime concerns. In practice, protocol endpoints should be treated as untrusted and routed through the same policy, approval, and logging controls as any other tool surface.

\section{Memory, State, and Observability}
\subsection{Retrieval memory is not workflow state}
Common beginner errors are to equate memory with vector retrieval or to treat longer context windows as a state model. Retrieval is for knowledge access; workflow coherence still requires explicit state objects and execution logs.

\subsection{Recommended layered design}
\begin{enumerate}[leftmargin=1.2em]
  \item session memory for short-term context,
  \item task state for typed workflow progression,
  \item organization knowledge for retrieval,
  \item immutable audit logs for replay and compliance.
\end{enumerate}

\subsection{Replay and debugging}
Long-horizon tasks need checkpoint, replay, and rollback by design. Observability should capture tool arguments, policy decisions, state transitions, and external side effects, not only final answers.

\subsection{Externalized context}
Connector-accessible systems increasingly hold part of the effective memory of the runtime. This improves capability, but it also turns access control, provenance, and audit into part of memory design.

\section{Training Paradigms for Agentic Reliability}
\subsection{Pre-train effects}
Pre-training still sets the ceiling for long-context robustness, tool-use priors, and code-edit competence.

\subsection{Mid-train environment shaping}
Interactive environments matter because they shape decomposition, recovery, and action discipline under delayed feedback.

\subsection{Post-train loops}
Different post-train methods target different failures:
\begin{itemize}[leftmargin=1.2em]
  \item SFT improves tool-call format and process compliance,
  \item RL or verifier-guided optimization targets long-horizon reliability,
  \item self-play and debate are useful only when backed by strict process-level evaluation.
\end{itemize}

\subsection{Practical interpretation}
In 2025--2026 production settings, post-train quality is inseparable from evaluation-loop quality. Many apparent ``agent'' gains come from better runtime design and evaluation, not from novel coordination topologies.

\section{Evaluation and Diagnostics}
\subsection{Why static scores are insufficient}
Outcome-only metrics hide process brittleness. Two systems with similar success rates can differ sharply in rollback frequency, unsafe tool calls, and cost variance.

\subsection{Four-layer evaluation model}
\begin{enumerate}[leftmargin=1.2em]
  \item outcome layer: success, cost, latency,
  \item process layer: failure taxonomy, tool misuse, retry cascades,
  \item safety layer: policy violations, injection success rate, leakage risk,
  \item behavior-realism layer: role consistency, convergence quality, and interaction dynamics for debate or social tasks.
\end{enumerate}

\subsection{SWE benchmark ecosystem}
SWE-bench Verified and Live matter because they move evaluation from static tasks toward realistic software-engineering distributions and contamination-resistant testing.

\subsection{Deployment gating}
Evaluation should be part of deployment gates. Every model, tool-policy, or prompt-template update should run benchmark regressions, include live tasks where feasible, and satisfy predefined process and safety thresholds.

\subsection{Why non-SWE coverage still matters}
Coding benchmarks are the strongest industrial stress test, but they do not cover coordination realism, role consistency, or deliberative behavior. Debate, collaborative QA, and opinion-dynamics lines remain useful when they supply benchmarks, datasets, or behavioral diagnostics.

\section{Infrastructure and Deployment}
\subsection{Runtime essentials}
Production runtime requirements are concrete:
\begin{itemize}[leftmargin=1.2em]
  \item explicit transitions,
  \item retry budgets and timeout policies,
  \item rollback paths,
  \item replay-capable tracing,
  \item budget-aware scheduling.
\end{itemize}

\subsection{Security and governance plane}
Once tools touch business systems, authn/authz, scoped credentials, secret isolation, approval gates, policy engines, and audit export are non-optional.

\subsection{Deployment venue is secondary}
Cloud vs self-hosted is not a core MAS design axis. It matters only when data residency, model control, integration constraints, or internal ops capacity change which runtime is feasible.

\section{Safety and Governance}
\subsection{Beyond prompt injection}
Once tools, protocols, and multiple agents are involved, the threat surface extends beyond the prompt layer:
\begin{itemize}[leftmargin=1.2em]
  \item tool schema injection,
  \item endpoint spoofing,
  \item connector supply-chain compromise,
  \item privilege escalation through chained tools.
\end{itemize}

\subsection{Control architecture}
Robust control combines least privilege, signed manifests, allowlists, pre-execution policy checks, audit trails, and continuous red-team evaluation. Better reasoning does not remove these requirements.

\section{High-Value Application Domains}
\subsection{Software engineering agents}
This is the strongest industrial pull and the best stress test because outputs are executable and verifiable. The hard parts are repository understanding, test-grounded edits, recovery after failed patches, and contamination-resistant evaluation.

\subsection{Enterprise workflow automation}
Ticketing, CRM/ERP workflows, compliance drafting, and multi-system operations are valuable when approvals, policy boundaries, and audit requirements are explicit.

\subsection{Research automation}
Literature triage, experiment planning, code replication, and reproducibility checks are promising but require strong citation integrity and version-control discipline.

\subsection{Collaborative QA and social-interaction tasks}
Collaborative QA, deliberation, and social-interaction tasks still matter because they expose coordination quality, role consistency, and realism properties that coding benchmarks do not measure.

\subsection{Tool-heavy desktop and browser automation}
Desktop and browser automation has clear economic pull, but only under strict permissioning, audit, and rollback controls because the action space is broad and often irreversible.

\section{Limits and Decision Boundary}
\subsection{Favorable conditions}
Multi-agent designs are justified when:
\begin{itemize}[leftmargin=1.2em]
  \item task decomposition is natural and stable,
  \item verifier independence improves reliability,
  \item parallel execution reduces wall-clock time,
  \item failures are isolatable at subtask boundaries.
\end{itemize}

\subsection{Unfavorable conditions}
Multi-agent is often counterproductive when:
\begin{itemize}[leftmargin=1.2em]
  \item the task is short and tool path is simple,
  \item observability is weak,
  \item coordination overhead dominates compute budget,
  \item team maturity is insufficient for policy + eval operations.
\end{itemize}

\section{Builder Path: Minimum Viable Stack for Shipping Teams}
\begin{enumerate}[leftmargin=1.2em]
  \item workflow orchestrator with typed state and explicit transitions,
  \item policy-mediated tool gateway with approval and audit hooks,
  \item structured state store plus immutable event log,
  \item benchmark and diagnostics harness covering both outcome and process,
  \item safety guardrails for injection, tool abuse, and connector compromise.
\end{enumerate}

\section{Research Path: Open Problems (2026Q1--Q4)}
\noindent The questions below are framed as the most important research watchpoints for the remainder of 2026, rather than a multi-year horizon.
\begin{itemize}[leftmargin=1.2em]
  \item durable semantic interoperability across vendor runtimes,
  \item standard protocol-security baselines for connector ecosystems,
  \item contamination-resistant live evaluation for non-SWE domains,
  \item formal guarantees for long-horizon workflow stability,
  \item cost-aware coordination policies under dynamic workloads.
\end{itemize}

\section{Conclusion}
The clearest 2025--2026Q1 trend is runtime maturation. Progress now depends more on protocol discipline, controllability, evaluation depth, and safety governance than on prompt design alone. Multi-agent architecture is justified only when it yields decomposition, independent checking, or parallel execution under constraints; otherwise it adds coordination cost. The field needs better systems design and behavior-aware evaluation, not more ungrounded role-play.

\section*{Evidence Map (Working Table)}
\begin{table*}[t]
\small
\centering
\begin{tabular}{>{\raggedright\arraybackslash}p{0.16\textwidth} >{\raggedright\arraybackslash}p{0.10\textwidth} >{\raggedright\arraybackslash}p{0.10\textwidth} >{\raggedright\arraybackslash}p{0.41\textwidth} >{\raggedright\arraybackslash}p{0.11\textwidth}}
\toprule
Artifact & Type & Tier & Why it matters & Industrial value \\
\midrule
AgentBench & Benchmark & A & Historical agent-eval baseline before the 2025 shift & Medium \\
SWE-bench Verified & Benchmark & A & Reliable coding-agent regression set with human validation & High \\
SWE-Gym & Paper/Env & A/B & Training and verifier environment for SWE agents & High \\
MAST / MAST-Data & Paper/Dataset & A/B & Failure taxonomy and trajectory-level diagnosis & Medium-High \\
SWE-bench Live & Benchmark & A & Anti-contamination, continuously refreshed SWE pressure test & High \\
DEBATE & Benchmark/Dataset & B & Opinion-dynamics and social-interaction realism coverage & Medium \\
M-MAD & Paper/Code & B & Debate-style coordination used as an evaluation system & Medium \\
CIR3 & Journal paper & B & Collaborative QA application coverage beyond SWE & Medium \\
MCP launch + spec & Protocol & A & Standardized model-tool-data interface layer & High \\
OpenAI platform docs & Doc & A & Runtime/tooling/tracing direction, deployability hooks & High \\
LangGraph & Framework & A & Graph orchestration with explicit workflow semantics & High \\
AutoGen & Framework & A & Multi-agent coordination plus memory/runtime evolution & High \\
LlamaIndex workflows & Framework/Doc & A & Knowledge-centric agent pipeline design & High \\
CrewAI & Framework & B & High-level role-oriented composition and comparison point & Medium-High \\
Sonar SWE-bench release & Industry signal & B & Benchmark-to-product translation in 2026 messaging & High \\
\bottomrule
\end{tabular}
\end{table*}

\begin{thebibliography}{99}
\bibitem{agentbench} AgentBench. \url{https://arxiv.org/abs/2308.03688}.
\bibitem{swebench_verified} Introducing SWE-bench Verified. \url{https://openai.com/index/introducing-swe-bench-verified/}.
\bibitem{swegym} Training Software Engineering Agents and Verifiers with SWE-Gym. \url{https://arxiv.org/abs/2412.21139}.
\bibitem{mast} Why Do Multi-Agent LLM Systems Fail? \url{https://arxiv.org/abs/2503.13657}.
\bibitem{swelive} SWE-bench Goes Live! \url{https://arxiv.org/abs/2505.23419}.
\bibitem{mmad} M-MAD. \url{https://aclanthology.org/2025.acl-long.351/}.
\bibitem{debate} DEBATE. \url{https://arxiv.org/abs/2510.25110}.
\bibitem{cir3} Coordinated LLM multi-agent systems for collaborative question-answer generation. \url{https://doi.org/10.1016/j.knosys.2025.114627}.
\bibitem{mcp_launch} Introducing the Model Context Protocol. \url{https://www.anthropic.com/news/model-context-protocol}.
\bibitem{openai_docs} OpenAI Platform Documentation. \url{https://platform.openai.com/docs}.
\bibitem{mcp} Model Context Protocol. \url{https://modelcontextprotocol.io/}.
\bibitem{autogen} Microsoft AutoGen. \url{https://github.com/microsoft/autogen}.
\bibitem{langgraph} LangGraph. \url{https://docs.langchain.com/oss/python/langgraph}.
\bibitem{llamaindex} LlamaIndex Multi-Agent Workflows. \url{https://docs.llamaindex.ai/en/stable/understanding/agent/multi_agent/}.
\bibitem{crewai} CrewAI. \url{https://github.com/crewAIInc/crewAI}.
\bibitem{sonar} Sonar Claims Top Spot on SWE-bench leaderboard. \url{https://www.sonarsource.com/company/press-releases/sonar-claims-top-spot-on-swe-bench-leaderboard/}.
\end{thebibliography}

\end{document}
